---
title: "Hw02-Explore Gapminder and use dplyr"
output: github_document
---
Load gapminder dataset and tidyverse and knitr packages.

```{r, message=FALSE}
library(gapminder)
library(tidyverse)
library(knitr)
```

## Small test of the data

* **Is it a data.frame, a matrix, a vector, a list?**

```{r}
gapminder
```

Look at the structure of gapminder

```{r}
str(gapminder)
```

Obviously, gapminder is a data frame. More specifically, it is a "tibble".

* **What is its class?**

```{r}
class(gapminder)
```

It seems like gapminder is a table and data frame.

* **How many variables/columns?**

```{r}
ncol(gapminder)
```

* **How many rows/observations?**

```{r}
nrow(gapminder)
```

* **Can you get these facts about “extent” or “size” in more than one way? Can you imagine different functions being useful in different contexts?**

`dim()` function can be used to get the number of rows and columns.

```{r}
dim(gapminder)
```

`summary()` and `str()` can help us get the content summary of gapminder. 

```{r}
summary(gapminder)
str(gapminder)
```

Sometimes we just want to filter one column of gapminder. For example, we want to see how many countries included in gapminder. 

```{r}
length(unique(gapminder$country))
```

* **What data type is each variable?**

```{r}
typeof(gapminder$country)
typeof(gapminder$continent)
typeof(gapminder$year)
typeof(gapminder$lifeExp)
typeof(gapminder$pop)
typeof(gapminder$gdpPercap)
```

## Explore individual variables

**Categorical variable: country and continent**

Let us look at possible values of continent. 

```{r}
unique(select(gapminder,continent))
```
5 continents are listed in the gapminder.

Then we will explore the number of countries in each continent.

```{r}
group_by(gapminder,continent)%>%
  summarise(numOfcountries=length(unique(country)))%>%
  ggplot(aes(x=continent,weight=numOfcountries))+geom_bar()+
  labs(title="The number of countries in each continent",x="continent",      y="number of countries")
```

Next, find out how many countries are listed?

```{r}
unique(select(gapminder,country))
```
142 countries are included in the gapminder.

Then let us explore how many entries for each country in this dataset. Take Canada as an example:

```{r}
filter(gapminder,country=="Canada") %>%
 select(country,year)
```

It shows that there are 12 entries for each country, every 5 years as an interval. 

**Quantitative variables: lifeExp, year, gdpPercap**

let's start with year.

```{r}
range(gapminder$year)
```

It shows that the data in the gapminder is from year 1952 to 2007.

Next we explore gdpPercap.

```{r}
summary(gapminder$gdpPercap)
```
The range is from 241.2 to 113523.1. It is a big range! 

Then we can make a table to demonstrate the gdpPercap summary in different continents from 1952 to 2007.

```{r}
group_by(gapminder,continent)%>%
  summarize(minValue=min(gdpPercap), 
            medianValue=median(gdpPercap),
            maxValue=max(gdpPercap))%>%kable()
```

In 2007, we can find out which country has highest gdpPercap?

```{r}
filter(gapminder,year==2007)%>%
  arrange(desc(gdpPercap))%>%
  select(country,gdpPercap)%>%
  head(10)
```

Norway has top gdpPercap in 2007!

Then we can graph gdpPercap in 2007 as a histogram.

```{r}
filter(gapminder,year==2007)%>%
  ggplot(aes(x=gdpPercap))+
  geom_histogram(bins=50)+
  labs(title="Histogram of gdpPercap in 2007")
```

Finally, we explore lifeExp variable.

```{r}
summary(gapminder$lifeExp)
```

like above, make a table to show mean values of lifeExp in 1952 and 2007 on different continents.

```{r}
filter(gapminder,year==1952|year==2007)%>%
  group_by(continent,year)%>%
  summarize(meanValue=mean(lifeExp))%>%
  kable()
```

We can find that mean value of lifeExp in each continent has a big increase.

Next, we will graph mean value of lifeExp in each continent from 1952 to 2007.

```{r}
group_by(gapminder,continent,year)%>%
  summarise(meanVal=mean(lifeExp))%>%
  ggplot(aes(x=year,y=meanVal,color=continent))+
  geom_point()+
  geom_smooth(se=FALSE)+
  labs(title="mean value of lifeExp in each continent")
```

## **Explore various plot types**

**Scatterplot**

let's begin with pop variable.

Firstly, we explore pop in each continent from 1952 to 2007.

```{r}
ggplot(gapminder,aes(x=continent,y=pop))+
  geom_point(alpha=0.2,color="blue")+
  stat_summary(fun.y=mean,color="red",geom = "point")+
  labs(title="population in each continent from 1952 to 2007")
```

Some countries in Asia has quite a big population.

Next, compare population of China and Canada.

```{r}
filter(gapminder,country=="China"|country=="Canada")%>%
  ggplot(aes(x=year,y=pop,color=country))+geom_point(alpha=0.5)+
  geom_smooth(se=FALSE)+
  labs(title="population of China and Canada")
```

That is amazing! The population difference between two countries is quite big.

Next variable we want to explore is gdpPercap vs year.

We want to graph gdpPercap vs year in every continent.

```{r}
ggplot(gapminder,aes(x=year,y=gdpPercap))+
  geom_point(color="blue")+
  stat_summary(fun.y=mean,color="red",geom = "point")+
  facet_wrap(~continent)+
  labs(title="gdpPercap vs year in each continent")
```

Then we can look at gdpPercap in top 10 European countries in 2007.

```{r}
filter(gapminder,year==2007,continent=="Europe")%>%
  arrange(desc(gdpPercap))%>%
  head(10)%>%
  select(country,gdpPercap)%>%
  ggplot(aes(x=country,y=gdpPercap))+
  geom_point(color="red")+
  labs(title="gdpPercap vs year in top 10 European countries in 2007")
```

**Histogram and density plot**

Let's see lifeExp distribution from 1952 to 2007.

```{r}
  ggplot(gapminder,aes(x=lifeExp,fill=continent))+
  geom_histogram()+
  labs(title="lifeExp distribution from 1952 to 2007")
```

Next look at the density plot of lifeExp distribution from 1952 to 2007.

```{r}
  ggplot(gapminder,aes(x=lifeExp,color=continent))+
  geom_density()+
  labs(title="lifeExp distribution from 1952 to 2007")
```

**Barplot**

let's see the total population of each continent in 2007.

```{r}
filter(gapminder,year==2007)%>%
  group_by(continent)%>%
  summarise(total = sum(as.numeric(pop)))%>%
  ggplot(aes(x=continent,weight=total))+
  geom_bar()+labs(title="total population of each continent in 2007",x="continent",y="population")
```

Let's see total gdp of each continent in 2007.

```{r}
e<-filter(gapminder,year==2007)%>%
   group_by(continent,country)%>%
   summarise(country_gdp=pop*gdpPercap)%>%
   group_by(continent)%>%
   summarise(total_gdp=sum(country_gdp))
ggplot(e,aes(x=continent,weight=total_gdp))+
  geom_bar()+
  labs(title="total gdp of each continent in 2007",x="continent",y="total gdp")
```

## **But I want to do more!**

```{r}
filter(gapminder,country==c("Rwanda", "Afghanistan"))%>%kable()
```

Obviously, this output is not correct, missing some information. 

There are two ways to correct it. 

First method:

```{r}
filter(gapminder,country=="Rwanda"|country=="Afghanistan")%>%kable()
```

Second method:

```{r}
filter(gapminder,country %in% c("Rwanda","Afghanistan"))%>%kable()
```

